What's in a name? Disambiguation strategies for relational descriptions

### 2023-05-19 2023 Disambiguation/Disambiguation.tex
Wrote a selection of disambiguation algorithms we can choose from.

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx,stmaryrd}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bbold,bbm}
\makeatletter
\def\amsbb{\use@mathgroup \M@U \symAMSb}
\makeatother
\usepackage[cal=boondoxo]{mathalfa}

@@

\begin{align}
\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} &\typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}} \\
\mathrm{before}_{\mathrm{Task * Task}} \typeinter \typeident_\mathrm{Task} &\typesubset \typenil_\mathrm{Task * Task}
\end{align}

The intended meaning behind these invariants is: 1) The set of tasks that needs to be performed before some task $x$, all need to be scheduled at times before the time at which $x$ itself is scheduled. 2) It makes no sense to require a task to be performed before itself.

@@

For the other expression, we cannot omit writing all types: $\mathrm{before} \typeinter \typeident \typesubset \typenil$ could be interpreted as:
$\mathrm{before}_{\mathrm{Time * Time}} \typeinter \typeident_{\mathrm{Time}} \typesubset \typenil_{\mathrm{Time * Time}}$.
To avoid this issue, one might write \emph{some} type information: $\mathrm{before} \typeinter \typeident \typesubset \typenil_{\mathrm{Task * Task}}$.
From this, it should become clear which relations were intended (namely $\mathrm{before}_\mathrm{Task * Task}$ and $\typeident_\mathrm{Task}$), even though they appear elsewhere.

@@

\section{Solving the disambiguation problem without generalization}
As an aside, we ask whether a compiler can be expected to figure out which relations are intended in reasonable time. In this section, we answer this question for a very simple setting in the affirmative by giving a simple efficient algorithm.

We can define the disambiguation problem as follows:

A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible signatures of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.

As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
It's (here unique) solution would be:
\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cup \left(\mathrm{scheduled}_{\mathrm{Task * Time}};\mathrm{before}_{\mathrm{Time * Time}}\right)\]

We can solve the disambiguation problem using two passes through $E$.
In the first pass, we decorate the expression tree for $E$ with all the possible types for each sub-expression bottom up.
Let $p(e)$ denote the possible types, then $p(n) = \mathcal{t}(n)$, and $p(e_1;e_2) = \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}$, and $p(e_1 \cup e_2) = p(e_1) \cap p(e_2)$.
If $p(E)$ is the empty set, then there is no solution to the disambiguation problem, and we are done.
If $p(E)$ contains multiple elements, then there are multiple solutions to the disambiguation problem.
We can arbitrarily select a single element from $p(E)$ to get an arbitrary solution\footnote{In practice, we might want to indicate that $E$ is ambiguous, which could require running the next step twice: once for two different elements in $p(E)$.}.

% @Bas, deze paragraaf is wellicht leesbaarder als je gewoon de functie à la Haskell uitschrijft.
% @Stef, zo staat het er al.
For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
Then $d(n,t) = n_{[t]}$, $d(e_1 \cup e_2, t) = d(e_1,t) \cup d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
Such a $z$ is guaranteed to exist by how $p$ is defined.

As an example, let's look at $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
Then $p(\mathrm{before}) =  \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ following $\mathcal{t}$ and similarly $p(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$.
This gives subterms $p(\mathrm{before};\mathrm{scheduled}) = p(\mathrm{scheduled};\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Time})\}$ so $p(E) = \{(\mathrm{Task}, \mathrm{Time})\}$.

Top-down, $d(E,t) = d(\mathrm{before};\mathrm{scheduled},t) \cup d(\mathrm{scheduled};\mathrm{before},t)$ with $t=(\mathrm{Task}, \mathrm{Time})$.
Since $E$ is not ambiguous, we have only one possible way to continue as we solve the disambiguation problem:
\begin{align*}d(\mathrm{before};\mathrm{scheduled},t) &= d(\mathrm{before},(\mathrm{Task}, \mathrm{Task})) ;  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time})) \\
&=\mathrm{before}_{\mathrm{Task} * \mathrm{Task}}; \mathrm{scheduled}_{\mathrm{Task}* \mathrm{Time}}\\
d(\mathrm{scheduled};\mathrm{before},t) &=  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time}));d(\mathrm{before},(\mathrm{Time}, \mathrm{Time}))\\
&=\mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}}; \mathrm{before}_{\mathrm{Time}* \mathrm{Time}}\\
\end{align*}

The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.
It can even be used to disambiguate special relations like the identity relation.
Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
The signatures are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.

\section{Solving Disambiguation with Generalization}
We first observe that the disambiguation algorithm discussed above fails for our running example.
Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
Now the apply the algorithm on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
Since the disambiguation algorithm does not consider sub-typing, it would conclude there are is no way to well-type this expression.
In this section we address this shortcoming.

@@

%This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
%Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
%% @Bas, de volgende zin snap ik niet...
%The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
%%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
%%In this paper, we relax that requirement.

@@

Let $\mathcal{S}$ be the language of heterogeneous relation algebra with subtyping.
It is defined by the following grammar, where $n\in\mathcal{N}$ and $A,B\in\mathcal{T}$:
\[
\varphi,\psi\in\mathcal{S} \mathbin{:=} n \mid \typetyped{\varphi}{A*B} \mid \typevee \mid \typeident \mid \typenil \mid \varphi \typeunion \psi \mid \varphi \typeinter \psi \mid \varphi \typesemi \psi \mid \typecomp{\psi}\mid \typeconv{\psi}
\]
The grammar $\mathcal{S}$ allows the programmer to write relations with their type, $\typetyped{n}{A*B}$ and $\typetyped{\typevee}{A*B}$, as usual.
However, the grammar is slightly more general, allowing type annotations for arbitrary terms $\typetyped{\varphi}{A*B}$.

Our aim is to disambiguate expressions written in this language into `normal' heterogeneous relation algebra, that is: without subtyping.
To signal certain type errors, we add an additional element $\bot$ to the set of concepts, that we will assume is different from every element in $\mathcal{T}$.
Since $\mathcal{T}$ is referred to as the set of concepts, we will refer to $\mathcal{T}\cup\{\bot\}$ as the set of pseudo-concepts.
We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}\cup\{\bot\}$ for all other occurrences of $A$ and $B$:
\[
\varphi,\psi\in\mathcal{R} \mathbin{:=} n_{A*B} \mid \epsilon_{A*B} \mid T_{A*B} \mid 1_{A} \mid F_{A*B} \mid \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi}
\]

The type of a term in $\mathcal{R}$ is given by a partial typing function, which we define inductively via the relation symbol $:$ as follows:
\begin{align*}
x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$}\\
1_{A} &: A * A&\\
\varphi \odot \psi &: A * B &\quad\text{For $\odot \in \{\cup,\cap\}$, $\varphi : A * B$ and $\psi : A * B$}\\
\varphi ; \psi &: A * B &\text{For $\varphi : A * C$ and $\psi : C * B$}\\
\overline{\psi} &: A * B &\text{For $\psi : A * B$}\\
\conv{\psi} &: A * B &\text{For $\psi : B * A$}
\end{align*}
We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for $A,B\in \mathcal{T}$ and every sub-term is well-typed.
By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is univalent.
In other words: well-typed terms in $\mathcal{R}$ have precisely one type, all of its sub-terms are well-typed, and $\bot$ does not occur as a pseudo-concept.

\subsection{Translation process}
Rather than translating terms in $\mathcal{S}$ into $\mathcal{R}$ directly, we create an intermediate language $\mathcal{D}$, which contains $\mathcal{R}$ (that is: $\mathcal{R} \subseteq \mathcal{D}$).
This will allow us to use a purely syntactic translation from $\mathcal{S}$ to $\mathcal{D}$, after which we run a disambiguation algorithm to turn it into a term in $\mathcal{R}$.
The language $\mathcal{D}$ is defined as follows, where $\xi \in \mathcal{R}$, $\Xi \subseteq \mathcal{R}$, and $A,B\in\mathcal{T}$:
\[
\varphi,\psi\in\mathcal{D} \mathbin{:=} \xi \mid \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi} \mid \typetyped{\varphi}{A*B} \mid \Xi
\]

@@

The translation from $\mathcal{S}$ to $\mathcal{D}$ will depend on the users's \verb=CLASSIFY= statements.
We write $A\preceq B$ to indicate that the pair $A,B$ is in the transitive closure of the pairs written \verb=CLASSIFY _ ISA _=, meaning that all elements in $A$ must occur in $B$ whenever $A\preceq B$.
Moreover, we let $\bot$ be a least element in this preorder: $\bot \preceq A$ for all $A$.
Since there are no classify statements containing $\bot$ (the user cannot write them), the least element in $\preceq$ is unique.

The translation $s : \mathcal{S} \to \mathcal{D}$ is as follows:
\begin{align*}
s(n) &= \{n_{A*B} \mid (A,B)\in\mathcal{t}(n)\}\\
s(\typetyped{\varphi}{A*B}) &= \typetyped{s(\varphi)}{A*B}\\
s(\typevee) &= \{T_{A*B} \mid A,B\in\mathcal{T}\}\\
s(\typenil) &= \{F_{A*B} \mid A,B\in\mathcal{T}\}\\
s(\typeident) &= \{1_{A} \mid A\in\mathcal{T}\}\\
s(\varphi \typeunion \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cup\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid A\preceq B \}$} \\
s(\varphi \typeinter \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cap\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \}$}\\
s(\varphi \typesemi \psi) &= s(\varphi);\varepsilon;s(\psi)&\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \vee A\preceq B \}$}\\
s(\typecomp{\varphi}) &= \overline{s(\varphi)}\\
s(\typeconv{\varphi}) &= \conv{s(\varphi)}
\end{align*}

@@

Translating a term from $\mathcal{D}$ to $\mathcal{R}$ can be thought of as a search problem:
For each $\Xi$-term in $\mathcal{D}$, we can substitute it for an element in $\Xi$ to get a term in $\mathcal{R}$.
Our choice should satisfy the following conditions:
\begin{itemize}
\item The chosen term should be well-typed.
\item Depending on the choices that remain after applying these criteria, we need to decide to pick an arbitrary one or raise an error to the user.
\item We prefer some terms over others. In particular, we prefer to use $\epsilon_{A*B}$ relations where $A = B$ over those where $A\neq B$.
\end{itemize}

@@

To illustrate why we prefer to use $\varepsilon_{A*B}$ relations where $A=B$, consider the expression: $(\mathrm{before} \typeinter \typeident) \typeunion \typetyped{\typenil}{\mathrm{Task * Task}}$. This expression is similar to one of the terms from our running example, but written as a proper $\mathcal{S}$-term.
Possible disambiguated terms include:
\begin{itemize}
\item $\varepsilon_T;(\varepsilon_T;\mathrm{before}_T;\varepsilon_T \cap \varepsilon_T;\typeident_\mathrm{Task};\varepsilon_T);\varepsilon_T \cup \varepsilon_T;F_T;\varepsilon_T$ with $T = \mathrm{Task*Task}$
\item
\item
\end{itemize}

the term $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typevee \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ with $\mathrm{Task}\preceq\mathrm{Event}$ would not disambiguate to the obvious term, since $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typetyped{\typevee}{\mathrm{Task}*\mathrm{Task}} \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ is also well-typed.
To deal with this, we prefer terms in which the $\epsilon$ relations have the same source and target.

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx,stmaryrd}
\usepackage{amsmath,amsfonts,amssymb,mathtools}
\usepackage{bbold,bbm}
\makeatletter
\def\amsbb{\use@mathgroup \M@U \symAMSb}
\makeatother
\usepackage[cal=boondoxo]{mathalfa}

@@

\begin{align}
\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} &\typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}} \label{eqn:schedule_sync}
 \\
\typecomp{\typeident_\mathrm{Task}} &\typesubset \typecomp{\left(\mathrm{before}_{\mathrm{Task * Task}} \typeinter \typeconv{{\mathrm{before}_{\mathrm{Task * Task}}}}\right)}
\label{eqn:before_antisym}
\end{align}

The operations used to describe these invariants have their typical set-theoretic interpretation.
Invariant~\ref{eqn:schedule_sync} intends to express: The set of tasks that needs to be performed before some task $x$, all need to be scheduled at times before the time at which $x$ itself is scheduled.
Invariant~\ref{eqn:before_antisym} intends to express: A task cannot be scheduled both before and after some other task.
It is the contrapositive of the antisymmetry property: ${\mathrm{before}_{\mathrm{Task * Task}}} \typeinter {\typeconv{{\mathrm{before}_{\mathrm{Task * Task}}}}} \typesubset \typeident_\mathrm{Task}$.
We focus on this somewhat clumsy formulation of this property to help explain the choices we can make in disambiguation algorithms.

@@

For the other expression, we cannot omit writing types. The expression $\typecomp\typeident \typesubset \typecomp{\left(\mathrm{before} \typeinter \typeconv{{\mathrm{before}}}\right)}$ is truly ambiguous, as it could be interpreted as:
$\typecomp{\typeident_{\mathrm{Time}}} \typesubset \typecomp{\left(\mathrm{before}_{\mathrm{Time * Time}} \typeinter \typeconv{\mathrm{before}_{\mathrm{Time * Time}}}\right)}$.
To avoid this issue, one might write \emph{some} type information: $\typecomp\typeident_\mathrm{Task} \typesubset \typecomp{\left(\mathrm{before} \typeinter \typeconv{{\mathrm{before}}}\right)}$.
From this, it should become clear which relations were intended (namely $\mathrm{before}_\mathrm{Task * Task}$), even though these appear elsewhere.

@@

\section{Solving Disambiguation}
We first observe that the disambiguation algorithm discussed above fails for our running example.
Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
Now the apply the algorithm on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
Since the disambiguation algorithm does not consider sub-typing, it would conclude there are is no way to well-type this expression.
In this section we address this shortcoming.

@@

%This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as sub-typing.
%Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
%% @Bas, de volgende zin snap ik niet...
%The semantics of the language with sub-typing will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
%%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a sub-typing relation, so $\epsilon_{A*B}$ would exist only if $A$ is a sub-type of $B$.
%%In this paper, we relax that requirement.

@@

Let $\mathcal{S}$ be the language of heterogeneous relation algebra with sub-typing.
It is defined by the following grammar, where $n\in\mathcal{N}$ and $A,B\in\mathcal{T}$:
\[
\varphi,\psi\in\mathcal{S} \mathbin{:=} n \mid \typetyped{\varphi}{A*B} \mid \typevee \mid \typeident \mid \typenil \mid \varphi \typeunion \psi \mid \varphi \typeinter \psi \mid \varphi \typesemi \psi \mid \typecomp{\psi}\mid \typeconv{\psi}
\]
The grammar $\mathcal{S}$ allows the programmer to write relations with their type, $\typetyped{n}{A*B}$ and $\typetyped{\typevee}{A*B}$, as usual.
However, the grammar is slightly more general, allowing type annotations for arbitrary terms $\typetyped{\varphi}{A*B}$.

Our aim is to disambiguate expressions written in this language into `normal' heterogeneous relation algebra, that is: without sub-typing.
%To signal certain type errors, we add an additional element $\bot$ to the set of concepts, that we will assume is different from every element in $\mathcal{T}$.
%Since $\mathcal{T}$ is referred to as the set of concepts, we will refer to $\mathcal{T}\cup\{\bot\}$ as the set of pseudo-concepts.
We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}%\cup\{\bot\}
$ for all other occurrences of $A$ and $B$:
\[
\varphi,\psi\in\mathcal{R} \mathbin{:=} n_{A*B} \mid \epsilon_{A*B} \mid T_{A*B} \mid 1_{A} \mid F_{A*B} \mid \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi}
\]

The type of a term in $\mathcal{R}$ is given by a partial typing function, which we define inductively via the relation symbol $:$ as follows:
\begin{align*}
x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$}\\
1_{A} &: A * A&\\
\varphi \odot \psi &: A * B &\quad\text{For $\odot \in \{\cup,\cap\}$, $\varphi : A * B$ and $\psi : A * B$}\\
\varphi ; \psi &: A * B &\text{For $\varphi : A * C$ and $\psi : C * B$}\\
\overline{\psi} &: A * B &\text{For $\psi : A * B$}\\
\conv{\psi} &: A * B &\text{For $\psi : B * A$}
\end{align*}
We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for $A,B\in \mathcal{T}$.
As a consequence of this definition all sub-terms of a well-typed term are well-typed.
Moreover, the relation $:$ is univalent (that is: well-typed terms in $\mathcal{R}$ have precisely one type).%, and $\bot$ does not occur as a pseudo-concept.

\subsection{Translation process}
Rather than translating terms in $\mathcal{S}$ into $\mathcal{R}$ directly, we create an intermediate language $\mathcal{D}$.
This will allow us to use a purely syntactic translation from $\mathcal{S}$ to $\mathcal{D}$, after which we run a disambiguation algorithm to turn it into a term in $\mathcal{R}$.
The language $\mathcal{D}$ is defined as follows, where $\Xi \subseteq \mathcal{R}$, and $A,B\in\mathcal{T}$:
\[
\varphi,\psi\in\mathcal{D} \mathbin{:=} \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi} \mid \typetyped{\varphi}{A*B} \mid \Xi
\]

@@

The translation from $\mathcal{S}$ to $\mathcal{D}$ will depend on the users's \verb=CLASSIFY= statements.
We write $A\preceq B$ to indicate that the pair $A,B$ is in the transitive closure of the pairs written \verb=CLASSIFY _ ISA _=, meaning that all elements in $A$ must occur in $B$ whenever $A\preceq B$.
%Moreover, we let $\bot$ be a least element in this preorder: $\bot \preceq A$ for all $A$.
%Since there are no classify statements containing $\bot$ (the user cannot write them), the least element in $\preceq$ is unique.

The translation $s : \mathcal{S} \to \mathcal{D}$ is as follows:
\begin{align*}
s(n) &= \{n_{A*B} \mid (A,B)\in\mathcal{t}(n)\}\\
s(\typetyped{\varphi}{A*B}) &= \typetyped{s(\varphi)}{A*B}\\
s(\typevee) &= \{T_{A*B} \mid A,B\in\mathcal{T}\}\\
s(\typenil) &= \{F_{A*B} \mid A,B\in\mathcal{T}\}\\
s(\typeident) &= \{1_{A} \mid A\in\mathcal{T}\}\\
s(\varphi \typeunion \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cup\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid A\preceq B \}$} \\
s(\varphi \typeinter \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cap\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid A\preceq B \}$}\\
s(\varphi \typesemi \psi) &= s(\varphi);\varepsilon;s(\psi)&\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \vee A\preceq B \}$}\\
s(\typecomp{\varphi}) &= \overline{s(\varphi)}\\
s(\typeconv{\varphi}) &= \conv{s(\varphi)}
\end{align*}

@@

\section{Solving the disambiguation problem}
Translating a term from $\mathcal{D}$ to $\mathcal{R}$ can be thought of as a search problem:
For each $\Xi$-term in $\mathcal{D}$, we can substitute it for an element in $\Xi$ to get a term in $\mathcal{R}$.
Our choice should satisfy the following conditions:
\begin{itemize}
\item The chosen term should be well-typed.
\item We prefer some terms over others. In particular, we prefer to use $\epsilon_{A*B}$ relations where $A = B$ over those where $A\neq B$.
\item Depending on the choices that remain after applying these criteria, we need to decide to pick an arbitrary one or raise an error to the user.
\end{itemize}

@@

To illustrate why we prefer to use $\epsilon_{A*B}$ relations where $A=B$, consider the expression: $\typecomp{\typeident} \typeunion \typecomp{\left(\typetyped{\mathrm{before}}{Task*Task} \typeinter \typeconv{\mathrm{before}}\right)}$. This expression is similar to Invariant~\ref{eqn:before_antisym} from our running example, but written as a proper $\mathcal{S}$-term.
Possible disambiguated terms include:
\begin{align}
\epsilon_{t_0};\overline{1_{\mathrm{Task}}};\epsilon_{t_0} &\cup \epsilon_{t_0};\overline{\left(\epsilon_{t_0};\typetyped{\mathrm{before}_{t_0}}{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_0}\label{eqn:disambGood}\\
\epsilon_{t_3};\overline{1_{\mathrm{Event}}};\epsilon_{t_3} &\cup \epsilon_{t_1};\overline{\left(\epsilon_{t_0};\typetyped{\mathrm{before}_{t_0}}{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_2}\label{eqn:disambBad}\\
\shortintertext{\flushright{ with ${t_0} = \mathrm{Task*Task}$, ${t_1} = \mathrm{Event*Task}$, ${t_2} = \mathrm{Task*Event}$, and ${t_3} = \mathrm{Event*Event}$}}\nonumber{}
\end{align}

The intended term is Term~\ref{eqn:disambGood} here.
Its meaning is different from that of Term~\ref{eqn:disambBad}, since the term $\epsilon_{t_3};\typecomp{1_{\mathrm{Event}}};\epsilon_{t_3}$ can contain pairs in $\mathrm{Event}$ that $\epsilon_{t_0};\typecomp{1_{\mathrm{Task}}};\epsilon_{t_0}$ does not.
We could argue that the user must specify more type information here, since the meanings for these terms differ.
However, the term can be interpreted as one that does not use sub-typing in any meaningful way.
Since the purpose of a disambiguation system is to reduce verbosity, we should prefer the intended term over Term~\ref{eqn:disambBad}.
A criterium that distinguishes the two terms is that the intended disambiguation uses $\epsilon_{t_0}$ throughout, which will be the preferred term if we prefer $\epsilon_{A*B}$'s that do not employ sub-typing (that is: those where $A=B$).

What if the user writes $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$, a term forces the use of sub-typing? Now possible disambiguated terms include:
\begin{align}
\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_3};\overline{\left(\epsilon_{t_1};\mathrm{before}_{t_0};\epsilon_{t_2} \cap \epsilon_{t_1};\conv{\mathrm{before}_{t_0}};\epsilon_{t_2} \right)};\epsilon_{t_3}\label{eqn:disambOption}\\
\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_1};\overline{\left(\epsilon_{t_0};\mathrm{before}_{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_2}\label{eqn:disambOld}\\
\shortintertext{\flushright{ with ${t_0} = \mathrm{Task*Task}$, ${t_1} = \mathrm{Event*Task}$, ${t_2} = \mathrm{Task*Event}$, and ${t_3} = \mathrm{Event*Event}$}}\nonumber{}
\end{align}

Observe that again, the two terms have different semantics, due to the complement being computed over a different type.
In previous publications on the Ampersand type system, where sub-typing did not play a role, Term~\ref{eqn:disambOld} would be chosen over Term~\ref{eqn:disambOption}.
The rational here is that types are computed bottom up.
However, we have observed that sub-typing around complements can be error-prone, and the Term~\ref{eqn:disambOption} might well have been intended.
We therefore argue that an error should be raised here.
The algorithm presented can be configured either way: accepting the term as Term~\ref{eqn:disambOld}, or throwing an error.
%As a consequence, we argue that it is desirable to require more information from the user here.
%terms like $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$ are

\section{Solving the disambiguation problem}
%We define the disambiguation problem as follows:
%
%A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible signatures of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
%A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.
%
%As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
%It's (here unique) solution would be:
%\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cup \left(\mathrm{scheduled}_{\mathrm{Task * Time}};\mathrm{before}_{\mathrm{Time * Time}}\right)\]

The question of whether a type-correct version of a term exists can be answered by a simple bottom-up calculation of the possible types for that term.
We define $p : \mathcal{D} \rightarrow \powerset{\left(\mathcal{T}\times \mathcal{T}\right)}$, for `possible types' by:
\begin{align*}
p(\Xi) &= \{t \mid r : t \text{ for $r\in\Xi$}\}\\
p(e_1; e_2) &= \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}\\
p(e_1\cup e_2) = p(e_1\cap e_2) &= \{(x,y) \mid (x,y) \in p(e_1) \wedge (x,y)\in p(e_2)\}\\
p(\overline{e}) &= p(e)\\
p(\typetyped{e}{t}) &= \begin{cases}\{t\}\quad\text{if $t\in p(e_1)$}\\\emptyset\quad\text{otherwise}\end{cases}\\
p(\conv{e}) &= \{(y,x) \mid (x,y)\in p(e)\}
\end{align*}

If $p(e) = \emptyset$ then $e$ has a type error, and a smallest sub-term $s$ of $e$ for which $p(s)=\emptyset$ can be used to give a meaningful type error.
We can compute $p(e)$ in time that is at most quadratic in terms of $\left|\mathcal{T}\times \mathcal{T}\right|$, so computing $p$ can be done in polynomial time in $\left|\mathcal{T}\right|$.
For Ampersand, $\left|\mathcal{T}\right|$ is at most linear in the size of the user's script.

Aside from type errors, we can use $p$ to find ambiguous terms:
If there is a sub-term $\Xi$ of $e$ that can be replaced by both $x\in \Xi$ and $y\in Xi$ for some $x \neq y$, then $p(e')$ and $p(e'')$ are nonempty for $e'$ and $e''$ equal to $e$ under the substitution of $\Xi$ for $\{x\}$ and $\{y\}$ respectively.
Furthermore, we can use $p$ to find out whether a negation can have multiple types.
